---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# htcrosstabs

<!-- badges: start -->
<!-- badges: end -->

The goal of htcrosstabs is to easily create printable crosstabs from raw data frames.

## Installation

You can install the development version of htcrosstabs like so:

``` r
devtools::install_github("HealthTree-Research-Team/htcrosstabs")
```

And you can load it like so:

``` r
library(htcrosstabs)
```

```{r, include = FALSE}
library(htcrosstabs)
```


## Creating a Crosstab

A simple crosstab can be created from a one-column data frame.

```{r, include = FALSE}
sport_options <- c("Basketball", "Soccer", "Football", "Hockey", "Baseball")

sports <- data.frame(
    sport = sample(sport_options, 200, replace = TRUE)
)
```

```{r}
head(sports, 5)
new_crosstab <- crosstab(sports)
```

A crosstab can group data into cohorts by providing a two column data frame, one for the data and one for the cohorts. Simply provide the name of the cohort column when creating it.

```{r, include = FALSE}
age_options <- c("Child", "Teen", "Adult")

# Example data
sports_by_age <- data.frame(
    sport = sample(sport_options, 200, replace = TRUE),
    age = sample(age_options, 200, replace = TRUE)
)
```

```{r}
head(sports_by_age, 5)
new_crosstab <- crosstab(sports_by_age, cohort_col_name = "age")
```

## Crosstab Types

Crosstabs can hold multiple data types:
- Numeric
- Categorical
- Likert-like
- Multi-response

The data type is automatically determined when you create the crosstab.

### Categorical Crosstabs

Categorical crosstabs are the default data type. Character values, logical values, and factors will all be counted as categorical.

```{r}
head(sports_by_age, 5) # Categorical
new_crosstab <- crosstab(sports_by_age, "age")
```

### Numeric Crosstabs

Numeric crosstabs are the default when the data is numbers.

```{r, include = FALSE}
length_by_species <- iris[, c("Petal.Length", "Species"), drop = FALSE]
names(length_by_species) <- c("petal length", "species")
```

```{r}
head(length_by_species, 5)
new_crosstab <- crosstab(length_by_species, "species")
```

### Likert-Like Crosstabs

Likert crosstabs are created when categorical data is provided, along with a named vector to map values to numbers.

```{r, include = FALSE}
like_options <- c("Likes", "Neither", "Dislikes")
licorice_by_age <- data.frame(
    opinion = sample(like_options, 200, replace = TRUE),
    age = sample(age_options, 200, replace = TRUE)
)
```

```{r}
head(licorice_by_age, 5)
likert_map <- c("Likes" = 1, "Neither" = 0, "Dislikes" = -1)

new_crosstab <- crosstab(licorice_by_age, "age", var_map = likert_map)
```

### Multi-Response Crosstabs

Multi-response crosstabs are created by providing a list-column allowing multiple answers per row.

```{r, include = FALSE}
allergy_options <- c("Nuts", "Eggs", "Milk", "Gluten")
school_options <- c("Brighton High", "Cottonwood High", "Olympus High", "Alta High")

get_allergy_sample <- function() {
    num_choices <- sample(1:4, 1)
    sample(allergy_options, num_choices, replace = FALSE)
}

allergies_by_school <- data.frame(
    allergies = I(replicate(200, get_allergy_sample(), simplify = FALSE)),
    school = sample(school_options, 200, replace = TRUE)
)
```

```{r}
head(allergies_by_school, 5)
head(allergies_by_school$allergies, 5)
new_crosstab <- crosstab(allergies_by_school, "school")
```

